Neural Connectome of the Ctenophore Statocyst

### 2025-04-15 manuscript/Manuscript.qmd
MS: edit methods section

### Specimen Preparation, Transmission Electron Microscopy, and Image Processing

Larvae of *Mnemiopsis leidyi* (5 days old) were cryofixed using a high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) and immediately transferred to liquid nitrogen for storage. The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica). Cryo-substitution was performed by gradually raising the temperature, and the samples were embedded in epoxy resin. Serial sections of 50 nm thickness were prepared using a Reichert Jung Ultracut E ultramicrotome and a 45º DiATOME diamond knife. To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the carbon film surface negatively charged. Section ribbons were collected on the prepared glass slides, slowly dried to ensure proper stretching, and firmly adhered to the glass surface. The sections on glass were stained with UranyLess and lead citrate (Reynolds) using airless staining bottles (Delta Microscopies). The glass slides were mounted on STEM-specific stages (Zeiss) using Copper Foil EMI Shielding Tape (3M). Imaging was performed using a Gemini SEM 500 (Zeiss) equipped with SmartSEM and Atlas 5 imaging software (Zeiss). The full dataset consisted of 620 serial sections, each 50 nm thick. The imaging resolution was 2.8 nm/pixel, using electron holography transmission at an acceleration voltage of 1.5 kV (in-lens detector, dwell time: 3 µs).

### Image-Stack Realignment and Spatial Data Transformation in CATMAID

@@

The alignment process consisted of three stages, each progressively refining the spatial accuracy:

Rigid Alignment Initially, a rigid alignment was performed with the following parameters: least squares mode (linear feature correspondences), encompassing the entire layer range with the first layer as the reference. Only visible images were used, without propagation. The alignment was executed with an initial Gaussian blur of 1.6 pixels, three steps per scale octave, a minimum image size of 512 pixels, and a maximum of 2048 pixels. Additional parameters included a feature descriptor size of 8, orientation bins of 8, and a closest ratio of 0.92. The alignment allowed clearing the cache, using 32 feature extraction threads, a maximal alignment error of 100 pixels, a minimal inlier ratio of 0.20, and a minimum of 12 inliers. The expected and desired transformations were set to rigid, with testing multiple hypotheses (tolerance: 5.00 pixels) and considering up to 5 neighboring layers, giving up after 5 failures. Regularization was performed with a maximal iteration of 1000, a maximal plateau width of 200, and a rigid lambda of 0.10.

Affine Alignment Following the rigid alignment, an affine alignment was applied using similar parameters, except the expected and desired transformations were set to affine. The minimal image size was reduced to 64 pixels, while the other parameters (e.g., Gaussian blur, feature descriptor size, inliers, and testing hypotheses) remained unchanged to ensure consistent processing.

Elastic Alignment Two iterations of elastic alignment were performed to fine-tune the spatial data. Key parameters included a block matching layer scale of 0.05, a search radius of 200 pixels, a block radius of 2000 pixels, and a resolution of 60. Correlation filters were employed with a minimal PMCC r of 0.10, a maximal curvature ratio of 1000, and a maximal second-best r/best r of 0.90. A local smoothness filter was applied with the approximate local transformation set to affine, a local region sigma of 1000 pixels, and an absolute maximal local displacement of 10 pixels (relative maximal displacement: 3.00). Pre-aligned layers were tested for up to 4 neighboring layers. The elastic alignment used a rigid approximation, maximal iterations of 3000, a plateau width of 200, spring mesh stiffness of 0.01, and a maximal stretch of 2000 pixels. A legacy optimizer was employed to enhance performance.

@@

The reconstruction of cells and connectivity was performed using CATMAID [@Saalfeld2009], analyzing a total of 619 serial sections. To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server [@Saalfeld2009; @Schneider-Mizell2016]. To mark the locations of cell bodies, we placed tags at the approximate center of each nucleus within the dataset. At each nuclear center, the radius of the single node was adjusted according to the size of the cell body in that specific layer. All skeletons were rooted at their respective cell bodies, and nodes were tagged as "soma." Synapses were identified based on four key structural features: the cell membrane, synaptic vesicles, the endoplasmic reticulum, and mitochondria. Most synapses could be verified across consecutive sections, ensuring accurate annotation and connectivity mapping.

@@

Each cell is assigned a specific type name based on its category (balancer, bridge, bristle, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types, or those that could not be classified: presence and number of cilia (biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)) in 4 ciliated cell types. Additionally, a quadrant number is appended to the cell type name following an underscore ("\_") to indicate the cell body’s location. For example, cells located in the first quadrant are annotated with "Q1." If a cell is situated between the first and second quadrants, the annotation "Q1Q2" is applied.

Cells of the same type within the same region are further distinguished by serial numbering. Each cell is also assigned multiple annotations, which can be utilized to query the database via CATMAID or the CATMAID API (e.g., using the R catmaid package). These annotations provide a structured and precise framework for identifying and analyzing specific cells, facilitating robust data integration and retrieval within the dataset.

This system ensures consistent and accurate tracking of cellular data, enabling comprehensive analysis of cellular structures and functions across regions.

### Imaging of Balancer Cilia activity

We conducted the analysis using the cydippid stage of *M. leidyi* at five days post-fertilization. A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass. Filtered natural seawater was introduced under the coverslip, along with the cy3dippids. By gently moving the coverslip, the desired orientation for observation was adjusted, and once positioned, slight pressure was applied to immobilize the cydippids. To stabilize the statolith's position, the microscope was tilted 90 degrees, arranging the stage vertically. The movement of the balancer cilia was observed using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS). Observations were made using a 40x objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27) with glycerine immersion. For recording, we employed a custom-made NIR LED strobe illumination system (wavelength: 850 nm) synchronized with the camera's exposure signals. The camera and LED strobe system were operated using the Video Capture Software BohNavi. Images were recorded at a resolution of 640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms pulses from the LED. To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph project in Fiji software.

@@

The following EM image stacks, including all traces and annotations, are available at https://catmaid.jekelylab.ex.ac.uk. These datasets can be queried either within CATMAID or through the CATMAID application programming interface (API) using the catmaid or natverse packages [@Bates2020]. The dataset encompasses all EM images (in JPG format), skeletons, meshes, node tags, connectors, and annotations. Additionally, we provide all R scripts used for data acquisition and figure generation (Jokura et al., 2025). All plots, figures (including anatomical renderings), and figure layouts can be fully reproduced using the provided R scripts. While the scripts are mostly organized by figure, some general-purpose scripts—for tasks such as loading libraries, accessing CATMAID data, and defining common functions—are shared across multiple figures.

>>>>

### Specimen Preparation, Volume Electron Microscopy, and Image Processing

Larvae of *Mnemiopsis leidyi* (five days old) were cryofixed using a high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) <!-- was this done in Exeter or Bergen? If Exeter, it is not the Baltec --> and immediately transferred to liquid nitrogen for storage. The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica). Cryo-substitution was carried out by gradually raising the temperature  <!-- steps? solutions? --> , and the samples were embedded in epoxy resin <!--type? -->. Serial untrithin sections of 50 nm thickness were prepared using a Reichert Jung Ultracut E ultramicrotome and a 45º DiATOME diamond knife <!-- wasn't it our Leica UC7? -->. To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the carbon film <!-- was there carbon film? --> surface negatively charged. Section ribbons were collected on the prepared glass slides, slowly dried to ensure proper stretching, and firmly adhered to the glass surface. The sections on glass were stained with UranyLess and lead citrate (Reynolds) using airless staining bottles (Delta Microscopies). The glass slides were mounted on STEM-specific stages (Zeiss) using Copper Foil EMI Shielding Tape (3M). Imaging was done on a Gemini SEM 500 (Zeiss) equipped with SmartSEM and Atlas 5 imaging software (Zeiss). The full dataset consisted of 620 serial sections, each 50 nm thick. The imaging resolution was 2.8 nm/pixel, using electron holography transmission <!-- is this true? --> at an acceleration voltage of 1.5 kV (in-lens detector, dwell time: 3 µs  <!-- sounds very low -->).

### Image-Stack Alignment and Export for CATMAID

@@

The alignment process consisted of three stages, each progressively refining the spatial accuracy (rigid, affine, elastic).

Initially, a rigid alignment was run with the following parameters: least squares mode (linear feature correspondences), encompassing the entire layer range with the first layer as the reference. Only visible images were used, without propagation. The alignment was executed with an initial Gaussian blur of 1.6 pixels, three steps per scale octave, a minimum image size of 512 pixels, and a maximum of 2048 pixels. Additional parameters included a feature descriptor size of 8, orientation bins of 8, and a closest ratio of 0.92. The alignment allowed clearing the cache, using 32 feature-extraction threads, a maximal alignment error of 100 pixels, a minimal inlier ratio of 0.20, and a minimum of 12 inliers. The expected and desired transformations were set to rigid, with testing multiple hypotheses (tolerance: 5.00 pixels) and considering up to 5 neighboring layers, giving up after 5 failures. Regularization was done with a maximal iteration of 1000, a maximal plateau width of 200, and a rigid lambda of 0.10.

Next we applied an affine alignment step with similar parameters, except the expected and desired transformations were set to affine. The minimal image size was reduced to 64 pixels, while the other parameters (Gaussian blur, feature descriptor size, inliers, and testing hypotheses) remained unchanged to ensure consistent processing.

Finally, we ran two iterations of elastic alignment to fine-tune the spatial data. Key parameters included a block-matching layer scale of 0.05, a search radius of 200 pixels, a block radius of 2000 pixels, and a resolution of 60. Correlation filters were employed with a minimal PMCC r of 0.10, a maximal curvature ratio of 1000, and a maximal second-best r/best r of 0.90. A local smoothness filter was applied with the approximate local transformation set to affine, a local region sigma of 1000 pixels, and an absolute maximal local displacement of 10 pixels (relative maximal displacement: 3.00). Pre-aligned layers were tested for up to 4 neighboring layers. The elastic alignment used a rigid approximation, maximal iterations of 3000, a plateau width of 200, spring mesh stiffness of 0.01, and a maximal stretch of 2000 pixels. A legacy optimizer was employed to enhance performance.

@@

For skeletonisation, annotation and tagging we used CATMAID installed on a local server [@Saalfeld2009; @Schneider-Mizell2016]. To mark the locations of cell bodies, we placed tags at the approximate center of each nucleus within the dataset. At each nuclear center, the radius of the single node was adjusted according to the size of the cell body in that specific layer. All skeletons were rooted at their respective cell bodies, and the root nodes were tagged as "soma." Synapses were identified based on four key structural features: the cell membrane, synaptic vesicles, the endoplasmic reticulum, and mitochondria. Most synapses could be verified across consecutive sections, ensuring accurate annotation and connectivity mapping.

@@

We assigned each cell a specific cell-type name based on its category (balancer, bridge, bristle, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types. Some cells only had more general features and were classified based on the presence/absence and number of cilia into four ciliated cell types:(biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)). Additionally, we appended the quadrant number to the cell type name after an underscore ("\_") to indicate the cell-body’s location. For example, cells located in the first quadrant are named with "_Q1." If a cell is situated between the first and second quadrants, we appended "Q1Q2" to the cell-type name.

We further distinguished cells of the same type within the same region by serial numbering. This way each cell in the volume has a unique name string. Each cell was also assigned multiple annotations, which can be utilized to query the database via CATMAID or the CATMAID API (e.g., using the R catmaid package [@Bates2020]). These annotations provide a structured and precise framework for identifying and analyzing specific cells, facilitating robust data integration and retrieval from the dataset.

### Imaging the Activity of Balancer Cilia

For ciliary imaging, we used the cydippid stage of *M. leidyi* at five days post-fertilization. We gently placed a coverslip with a thin layer of Vaseline applied to its two edges on a slide glass. Filtered natural seawater was introduced under the coverslip, along with the cydippids. By gently moving the coverslip, we could adjust the orientation of the larva, and once positioned, we applied slight pressure to immobilize the cydippids. We imaged the movement of balancer cilia with a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS) and a 40x glycerine-immersion objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27). To stabilize the statolith's position, we tilted the microscope 90 degrees, arranging the stage vertically. For recording, we employed a custom-made NIR LED strobe illumination system (wavelength: 850 nm) synchronized with the camera's exposure signals. The camera and LED strobe system were operated using the Video Capture Software BohNavi. Images were recorded at a resolution of 640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms pulses from the LED. To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph plugin in Fiji.

@@

The EM image stacks, including all traces and annotations, are available at https://catmaid.jekelylab.ex.ac.uk. The dataset encompasses all EM images (in JPG format), skeletons, meshes, node tags, connectors, and annotations. Additionally, we provide all R scripts used for data acquisition and figure generation (Jokura et al., 2025). All plots, figures (including anatomical renderings), and figure layouts can be fully reproduced using the provided R scripts. While the scripts are mostly organized by figure, some general-purpose scripts—for tasks such as loading libraries, accessing CATMAID data, and defining common functions—are shared across multiple figures.
